---
title: "DATA 607 HW week 5"
author: "Sergio Ortega Cruz"
date: "September 23, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment - Tidying and Transforming Data

The chart above describes arrival delays for two airlines across five destinations. Your task is to:

(1) Create a .CSV file (or optionally, a MySQL database!) that includes all of the information above. 

You're encouraged to use a "wide" structure similar to how the information appears above, so that you can practice tidying and transformations as described below.

(2) Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy and transform your data.

(3) Perform analysis to compare the arrival delays for the two airlines.

(4) Your code should be in an R Markdown file, posted to rpubs.com, and should include narrative descriptions of your data cleanup work, analysis, and conclusions. Please include in your homework submission:

The URL to the .Rmd file in your GitHub repository. and
The URL for your rpubs.com web page.



```{r db}

library(tidyr)
library(dplyr)
library(stringr)

flights <- read.csv("https://raw.githubusercontent.com/sortega7878/DATA607W5/master/flights.csv", 
                    sep = ",", header = TRUE)
flights
```
Data Cleaning and going Tidy.

At this point the initial data is tidy. Each variable (airline, city, delayed, ontime) is in its own column and each observation is in its own row.
```{r wrangling}

# Tidy up data, so that every variable is in a column and every observation is in a row

flights <- rename(flights, airline = ï..airline)
flights <- flights %>% 
  gather(city, number, 3:length(flights)) %>%   # Transfer 'city' variable into a column
  spread(status, number)                         # Split 'delayed'/'on time' variables into 2 columns

# Adjust city names (get rid of a period)
flights$city <- str_replace_all(flights$city, "\\.", " ")

# Remove space in the 'on time' variable name
flights <- rename(flights, ontime = `on time`)
flights
```

DATA ANALYSIS

Getting the ratio of delayed flights for all airlines and cities. We'll sort the data by delay starting with the worst airline/city.

```{r DA}

flights <- flights %>% 
  mutate(flights = delayed + ontime, delayed_ratio = delayed / flights) %>% 
  arrange(desc(delayed_ratio))
flights

```
Let us look compare how delays compare against each other using various criteria.

```{r avg}
# Average delay across all cities and airlines
flights %>% summarise(mean(delayed_ratio))


delay_city <- flights %>% 
  group_by(city) %>% 
  summarise(avg_delay = mean(delayed_ratio)) %>% 
  arrange(desc(avg_delay))
delay_city

```

```{r byAirline}

delay_airline <- flights %>% 
  group_by(airline) %>% 
  summarise(avg_delay = mean(delayed_ratio)) %>% 
  arrange(desc(avg_delay))
delay_airline

```
Map delays per city and per airline.

```{r graph}
library(ggplot2)

ggplot(flights, aes(x = city, y = delayed_ratio, group = airline, color = airline, shape = airline)) + 
  geom_point(aes(shape = airline)) + 
  geom_line() + 
  labs(x = "City", y = "Delay") + 
  theme(legend.title=element_blank())

```
Is important to note how the trends of delay matches the airport , knowing more data will help us determine why Alaska has bigger delay periods and then will be able to make corrective actions.



